Assessment of the KLGO Coastal Waterbird Survey Monitoring Program

### 2019-04-12 analysis/paper/paper.Rmd
In process of simplifying data frames based on the one created in data_wrangling. Have for the time being moved all data frame code to data_wrangling, which will be moved back to respective scripts after. Not done yet.
Change reference grid on Survey_Duration, minor updates on text, begin developing sequencing of target species.

```
# Things To Discuss with Madeleine
- add all libraries used in scripts to 'setup' code chunk (just above) rather than just loading in script so that it is easy to see what needs to be loaded before you can successfully knit.
- documentation in scripts

@@

# Issues to Resolve with content (narrative, figures, analyses)
- **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.).
- **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year. E.g., separate out any filtering process stemming from survey effort levels and timing. (thinking some thing like showing open circles for survey events PRIOR to first sighting...)
- Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on. E.g., those only seen <3 or 5 years??? Look at phenology graphs and determine threshold to apply. Consider some species N-ward and S-ward migration routes may differ.
- What is up with 2008 surveys? why such distinct timing?
- Think about Covariate EDA, if any...change in observers?
- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.

@@

Assessing a monitoring program focuses on four broad topics following the major Phases of a monitoring program (ibid):  
1. _why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
2. _how_ - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives? Given interim learning regarding system characteristics and improvements in measurement technologies, are the information goals (including accuracy and precision goals), actually feasible? If so, can the data collection be made more efficient, reducing the observation burden on staff resources?  
3. _how_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Do the data analysis methods properly account for impactful features of the data collection design? Are more informative analyses available given interim improvements in methods?  
4. _how_ - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?

@@

@@

> MADELEINE: With all these initial round of figures, don't get too caught up in refining fine details as we'll likely iterate a couple times and very well may end up deciding a totally different figure is more effective.  For example, this initial round of requests is a lot of dotplots, more or less, but we may end up realizing it is better to look at a summary of '% success per week' or something else that is a more continuous outcome variable. Hopefully some of that makes sense.

> SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, perhaps broken in to 1 or 2 hour blocks , y axis is ....this is a tough one so totally open to your ideas. How about y axis is % [0, 1.0] and for each year there is a different curve showing the % of planned surveys (as defined above) that were successfully completed that season within that time block; facet by site. So we are seeing whethere there are consistently problematic times of day at each site.

*Survey Effort*
Would like to get a sense of the consistency of (alternatively, variation in) survey effort (e.g., duration at each site) across Days of season and years.Like many of these figure requests, we are basically looking to check that there isn't anything glaring going on underneath so we are comfortable later summarizing across this particular variable, or making sure we account for it via some sort of weighting of observations (perhaps by focusing on a variant of something known in fisheries as Catch Per Unit Effort).

> MADELEINE: (that really isn't meant to come across as yelling, just trying to consistently label these so you can easily search for them) This is a very rough thought - feel free to suggest something else. We want to look at variation in 'duration' at each site across days of season and years. Maybe facet by season week, y axis is year, x axis is duration, horizontal boxplot of duration. This lumps all sites - maybe redo that graph again, once for each site?

@@

```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap=""}
print(plot_duration)
```

@@

*Survey Completion*

@@

Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).  **MADELEINE: Dig in to Site 1 to make sure there aren't messages hidden in 'notes' since that is the location that consists of cruise ship docking areas for a good chunk of the space.**
```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="Need to add caption.",warning=FALSE}

@@

#edit to use the name of the new plot
print(Plot1)
```

>>>>

```
# NOTES To Madeleine
- see edits in survey_duration.R to aid interpretation by the viewer (e.g., changes in x axis tick locations, etc.)
- documentation in scripts

@@

# Issues to Resolve with content (narrative, figures, analyses)
- **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year. E.g., separate out any filtering process stemming from survey effort levels and timing. (thinking some thing like showing open circles for survey events PRIOR to first sighting...)
- Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on. E.g., those only seen <3 or 5 years??? Look at phenology graphs and determine threshold to apply. Consider some species N-ward and S-ward migration routes may differ.
- What is up with 2008 surveys? why such distinct timing?
- Think about Covariate EDA, if any...change in observers?
- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.

@@

Assessing a monitoring program focuses on four broad topics following the major Phases of a monitoring program (ibid):  
1. _Why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
2. _How_ - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives? Given interim learning regarding system characteristics and improvements in measurement technologies, are the information goals (including accuracy and precision goals), actually feasible? If so, can the data collection be made more efficient, reducing the observation burden on staff resources?  
3. _How_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Do the data analysis methods properly account for impactful features of the data collection design? Are more informative analyses available given interim improvements in methods?  
4. _How_ - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?

@@

> **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.).

@@

> SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, perhaps broken in to 1 or 2 hour blocks , y axis is ....this is a tough one so totally open to your ideas. How about y axis is % [0, 1.0] and for each year there is a different curve showing the % of planned surveys (as defined above) that were successfully completed that season within that time block; facet by site. So we are seeing whethere there are consistently problematic times of day at each site. 
```{r SiteAccess, eval=TRUE, cache = FALSE, include = FALSE}
# STILL NEED TO CREATE
# source("../scripts/survey_access.R", echo=FALSE)
```

```{r Fig-SiteAccess, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="",warning= FALSE}
# print(plot_access)
```

*Survey Effort*
The time spent at each site during a survey has varied widely across the years, especially at Sites 1, 7, and 8 (Figure \@ref(fig:Fig-SurveyEffort)).  This variation in survey effort may be due to changes in numbers of birds occurring (and hence needing to be ID'd and recorded), variation among observers in identification skills, changes in visibility and area available for consistent viewing at each site, as well as potential changes in protocol across the years.  This variation in survey effort should be considered in any analysis of changes across years in occurrence or frequency of a species, or richness.  Depending on the analysis, this might best be done by adjusting the response variable from 'counts' to 'counts per unit effort', weighting observations by survey effort, or considering how survey duration might be accounted for as a covariate.

@@

```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="",warning= FALSE}
print(plot_duration)
```

@@

*Survey Completion*
> Ultimately, re-order these sections to flow better. Completion then Duration, etc., etc.

@@

Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).

**MADELEINE: Dig in to Site 1 to make sure there aren't messages hidden in 'notes' since that is the location that consists of cruise ship docking areas for a good chunk of the space.**

```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="Need to add caption.",warning=FALSE}

@@

#edit to use the name of the new plot
#print(Plot1)
```
